---
title: "FM Housing Price Prediction Project"
author: "Divine Gatebuka Iradukunda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ranger)
library(vetiver)
```

# Project Overview

This project uses Fargo-Moorhead housing data to develop a model that predicts 
the sold price of residential properties. The project includes data preparation, 
exploratory analysis, predictive modeling, model evaluation, and deployment 
through a Vetiver API and Shiny application.

## Data

The housing data for this project was provided through the DATA 470 course 
repository by Dr Gregory Tanner.

```{r load-data}
housing <- read_csv("FM_Housing_2018_2022_clean.csv")
glimpse(housing)
```

## Exploratory Data Analysis

Before building the predictive model, I examined the structure of the housing 
data and the distribution of the target variable, Sold Price.

```{r data-summary}
summary(housing$`Sold Price`)
```


```{r missing-values}
housing |>
  summarise(
    across(
      everything(),
      ~ sum(is.na(.))
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing"
  ) |>
  arrange(desc(missing))
```


```{r sold-price-distribution}
ggplot(housing, aes(x = `Sold Price`)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Distribution of Sold Prices",
    x = "Sold Price",
    y = "Number of Properties"
  )
```


## Interpretation

The distribution of Sold Price is right-skewed, with most properties selling at 
lower prices and a smaller number of properties selling for much higher prices. 
The long right tail shows that there are some high-priced properties in the dataset, 
with the maximum sold price being much higher than the typical property price.

This suggests that the model may need to account for the skewed distribution of 
housing prices when making predictions.

## Data Preparation

Before building the predictive model, I selected variables that could reasonably 
be used to predict a property's sold price. I excluded variables that would reveal 
information about the sale.


```{r select-variables}
model_data <- housing |>
  select(
    `Sold Price`,
    `List Price`,
    `Original List Price`,
    `Total SqFt.`,
    `Year Built`,
    `Total Bedrooms`,
    `Total Bathrooms`,
    `Garage Stalls`,
    `Lot Size SqFt`,
    `Above Grade Finished Area`,
    `Below Grade Finished Area`,
    `Above Grade Unfinished Area`,
    `Below Grade Unfinished Area`,
    `Property Type`,
    `Style`,
    `City`,
    `County`,
    `Garage Type`
  ) |>
  rename(
    sold_price = `Sold Price`,
    list_price = `List Price`,
    original_list_price = `Original List Price`,
    total_sqft = `Total SqFt.`,
    year_built = `Year Built`,
    total_bedrooms = `Total Bedrooms`,
    total_bathrooms = `Total Bathrooms`,
    garage_stalls = `Garage Stalls`,
    lot_size_sqft = `Lot Size SqFt`,
    above_grade_finished = `Above Grade Finished Area`,
    below_grade_finished = `Below Grade Finished Area`,
    above_grade_unfinished = `Above Grade Unfinished Area`,
    below_grade_unfinished = `Below Grade Unfinished Area`,
    property_type = `Property Type`,
    style = Style,
    city = City,
    county = County,
    garage_type = `Garage Type`
  )

glimpse(model_data)
```


### Why these?

We're keeping a mix of:

**Property characteristics**
- Total square footage
- Year built
- Bedrooms
- Bathrooms
- Garage
- Lot size
- Finished/unfinished areas

**Location**
- City
- County
- Property type
- Style

**Listing information**
- List Price
- Original List Price

This gives the model information that could reasonably be available when the 
property is being listed.

I excluded `Sold Price Per/SqFt` because it is calculated using the sold price 
and would cause target leakage. I also excluded variables such as `Sold Date` 
and `Days on Market` because they describe information that becomes available 
during or after the selling process rather than at the time of listing. 


## Training/Testing Data
```{r train-test-split}
set.seed(123)

train_index <- sample(
  seq_len(nrow(model_data)),
  size = 0.8 * nrow(model_data)
)

train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

nrow(train_data)
nrow(test_data)
```


I split the prepared data into training and testing sets, using 80% of the observations 
for model training and 20% for evaluating model performance/testing. A random seed was 
used so that the same split can be reproduced.


## Model Development

I used a Random Forest regression model to predict the sold price of properties. 
Random Forest was selected because it can capture nonlinear relationships between
housing characteristics and sold price while working with both numeric and categorical variables.

```{r random-forest}
set.seed(123)
model_data <- housing |>
  select(
    `Sold Price`,
    `List Price`,
    `Original List Price`,
    `Total SqFt.`,
    `Year Built`,
    `Total Bedrooms`,
    `Total Bathrooms`,
    `Garage Stalls`,
    `Lot Size SqFt`,
    `Above Grade Finished Area`,
    `Below Grade Finished Area`,
    `Above Grade Unfinished Area`,
    `Below Grade Unfinished Area`,
    `Property Type`,
    `Style`,
    `City`,
    `County`,
    `Garage Type`
  ) |>
  rename(
    sold_price = `Sold Price`,
    list_price = `List Price`,
    original_list_price = `Original List Price`,
    total_sqft = `Total SqFt.`,
    year_built = `Year Built`,
    total_bedrooms = `Total Bedrooms`,
    total_bathrooms = `Total Bathrooms`,
    garage_stalls = `Garage Stalls`,
    lot_size_sqft = `Lot Size SqFt`,
    above_grade_finished = `Above Grade Finished Area`,
    below_grade_finished = `Below Grade Finished Area`,
    above_grade_unfinished = `Above Grade Unfinished Area`,
    below_grade_unfinished = `Below Grade Unfinished Area`,
    property_type = `Property Type`,
    style = Style,
    city = City,
    county = County,
    garage_type = `Garage Type`
  )

glimpse(model_data)

# Random Forest Model
rf_model <- ranger(
  sold_price ~ .,
  data = train_data,
  num.trees = 500,
  importance = "impurity"
)

rf_model
```

## Model Evaluation

After training the Random Forest model, I evaluated its predictions using the testing 
dataset, which was not used during model training. 

```{r model-predictions}
predictions <- predict(
  rf_model,
  data = test_data
)$predictions

head(predictions)


actual <- test_data$sold_price
predicted <- predictions

rmse <- sqrt(mean((actual - predicted)^2))

mae <- mean(abs(actual - predicted))

r_squared <- 1 - sum((actual - predicted)^2) /
  sum((actual - mean(actual))^2)

model_performance <- tibble(
  Metric = c("RMSE", "MAE", "R-squared"),
  Value = c(rmse, mae, r_squared)
)

model_performance
```

### Interpretation

The Random Forest model performed well on the testing data, with an R-squared value of 0.934. 
This means that the model explains approximately 93.4% of the variation in sold prices for 
the properties in the testing set. 

The model had a mean absolute error (MAE) of approximately $8,126, meaning the average absolute 
difference between the predicted and actual sold prices was about $8,126. 

The RMSE was approximately $35,380. This means the model's prediction errors were around 
$35,380 on average, with larger errors having a bigger effect on the RMSE.

Overall, the results indicate that the model was able to predict sold prices reasonably well
on properties it had not been trained on.

### Visualization

```{r actual-vs-predicted}
prediction_results <- test_data |>
  select(sold_price) |>
  mutate(
    predicted_price = predictions
  )

ggplot(prediction_results, aes(x = sold_price, y = predicted_price)) +
  geom_point(alpha = 0.4) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  labs(
    title = "Actual vs. Predicted Sold Prices",
    x = "Actual Sold Price",
    y = "Predicted Sold Price"
  )
```
### Interpretation

Overall, the model's predictions were fairly close to the actual prices for most homes, 
especially those selling below $1 million. However, the model had larger errors for some 
of the most expensive homes and underestimated their prices. 

For example, one property sold for over $3 million, while the model predicted a price below 
$1.5 million.

### Vetiver Model

```{r vetiver-model}
v <- vetiver_model(
  rf_model,
  model_name = "fm-housing-price-model",
  prototype_data = train_data |>
    select(-sold_price) |>
    slice(1)
)

v
```
```{r save-vetiver-model}
saveRDS(v, "fm-housing-vetiver.rds")
```


